---
title: "Impact of Storms and Other Severe Weather Events On Both Public Health and Economic Problems"  
output: html_document
---

The basic goal of this report is to explore the NOAA Storm Database and answer the following questions  

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?  

A copy of the data file can be found at https://github.com/fissio/RepData_PeerAssessment2/blob/master/repdata-data-StormData.csv.bz2  

## Data Processing
### Read the storm data
```{r}
library(ggplot2)
library(grid)
library(gridExtra)
stormdata <- read.csv("repdata-data-StormData.csv")
```

### Analyse the data distribution over the years
```{r}
year <- as.numeric(format(as.Date(stormdata$BGN_DATE, format = "%m/%d/%Y %H:%M:%S"), "%Y"))
stormdata <- cbind(stormdata, year)
hist(stormdata$year, breaks = 60, xlab = "Year", main = "Distribution of Storm Data Over 1950 - 2010", col = "antiquewhite3")
```
#### In the earlier years of the database there are generally fewer events recorded,therefore we will only consider data collected from 1990 onwards. We will also reduce the data to include only columns that are relevant for this analysis
```{r}
stormdata <- subset(stormdata, stormdata$year >= 1990)
stormdata <- stormdata[,c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]

#Process data to determine what are the events that adversely affect population health
injury_results <- aggregate(stormdata[c("INJURIES")], list(evtype = stormdata$EVTYPE), sum)
fatalities_results <- aggregate(stormdata[c("FATALITIES")], list(evtype = stormdata$EVTYPE), sum)
combined_results <- merge(fatalities_results, injury_results, by.x = "evtype", by.y = "evtype")
combined_results$sum <- combined_results$FATALITIES + combined_results$INJURIES
injury_plot <- combined_results[order(-combined_results$INJURIES), ]
injury_plot <- head(injury_plot,10)
injury_plot$evtype <- factor(injury_plot$evtype, levels = injury_plot$evtype)
fatalities_plot <- combined_results[order(-combined_results$FATALITIES), ]
fatalities_plot <- head(fatalities_plot,10)
fatalities_plot$evtype <- factor(fatalities_plot$evtype, levels = fatalities_plot$evtype)
combined_plot <- combined_results[order(-combined_results$sum), ]
combined_plot <- head(combined_plot,10)
combined_plot$evtype <- factor(combined_plot$evtype, levels = combined_plot$evtype)
```

```{r}
#Process data to determine
unique(stormdata$PROPDMGEXP)
unique(stormdata$CROPDMGEXP)
```
#### There are undefined characters contained within the exponent of property and crop damage data. We will arbitrarily set them to zeros. We will also change exponent represented in their character representation to their numeric values.

```{r}
stormdata$PROPDMGEXP <- as.character(stormdata$PROPDMGEXP)
stormdata$PROPDMGEXP = gsub("\\-|\\+|\\?|h|H","0",stormdata$PROPDMGEXP)
stormdata$PROPDMGEXP = gsub("k|K", "1000", stormdata$PROPDMGEXP)
stormdata$PROPDMGEXP = gsub("m|M", "1000000", stormdata$PROPDMGEXP)
stormdata$PROPDMGEXP = gsub("b|B", "1000000000", stormdata$PROPDMGEXP)
stormdata$PROPDMGEXP <- as.numeric(stormdata$PROPDMGEXP)
stormdata$PROPDMGEXP[is.na(stormdata$PROPDMGEXP)] = 0

stormdata$CROPDMGEXP <- as.character(stormdata$CROPDMGEXP)
stormdata$CROPDMGEXP = gsub("\\?","0",stormdata$CROPDMGEXP)
stormdata$CROPDMGEXP = gsub("k|K", "1000", stormdata$CROPDMGEXP)
stormdata$CROPDMGEXP = gsub("m|M", "1000000", stormdata$CROPDMGEXP)
stormdata$CROPDMGEXP = gsub("b|B", "1000000000", stormdata$CROPDMGEXP)
stormdata$CROPDMGEXP <- as.numeric(stormdata$CROPDMGEXP)
stormdata$CROPDMGEXP[is.na(stormdata$CROPDMGEXP)] = 0

stormdata$PROPDMG <- stormdata$PROPDMG * stormdata$PROPDMGEXP
stormdata$CROPDMG <- stormdata$CROPDMG * stormdata$CROPDMGEXP

propdmg_results <- aggregate(stormdata[c("PROPDMG")], list(evtype = stormdata$EVTYPE), sum)
cropdmg_results <- aggregate(stormdata[c("CROPDMG")], list(evtype = stormdata$EVTYPE), sum)
propcropdmg_results <- merge(propdmg_results, cropdmg_results, by.x = "evtype", by.y = "evtype")
propcropdmg_results$sum <- propcropdmg_results$CROPDMG + propcropdmg_results$PROPDMG

cropdmg_plot <- propcropdmg_results[order(-propcropdmg_results$CROPDMG), ]
cropdmg_plot <- head(cropdmg_plot,10)
cropdmg_plot$evtype <- factor(cropdmg_plot$evtype, levels = cropdmg_plot$evtype)
propdmg_plot <- propcropdmg_results[order(-propcropdmg_results$PROPDMG), ]
propdmg_plot <- head(propdmg_plot,10)
propdmg_plot$evtype <- factor(propdmg_plot$evtype, levels = propdmg_plot$evtype)
propcropdmg_plot <- propcropdmg_results[order(-propcropdmg_results$sum), ]
propcropdmg_plot <- head(propcropdmg_plot,10)
propcropdmg_plot$evtype <- factor(propcropdmg_plot$evtype, levels = propcropdmg_plot$evtype)
```

## Results  

### Events Harmful to Population Health  
```{r}
plot_1 <- ggplot(injury_plot, aes(x=evtype, y=INJURIES)) + 
  geom_bar(stat = "identity", fill = "blue") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) + 
  ylab("Count") + ggtitle("Injuries") + 
  xlab("Event Type")
plot_2 <- ggplot(fatalities_plot, aes(x=evtype, y=FATALITIES)) + 
  geom_bar(stat = "identity", fill = "green") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) + 
  ylab("Count") + ggtitle("Fatalities") + 
  xlab("Event Type")
plot_3 <- ggplot(combined_plot, aes(x=evtype, y=sum)) + 
  geom_bar(stat = "identity", fill = "red") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) + 
  ylab("Count") + ggtitle("Fatalities+Injuries") + 
  xlab("Event Type")
grid.arrange(plot_1, plot_2, plot_3, ncol = 3, main = "Top 10 Events Affecting Population Health")
```




#### The top 3 events that resulted in most injuries are  
1. Tornado
2. Flood
3. Excessive Heat  

#### The top 3 events that resulted in most fatalities are
1. Excessive Heat
2. Tornado
3. Flash Flood

#### The top 3 events taking into consideration both injuries and fatalities that are most harmful to population health are  
1. Tornado
2. Excessive Heat
3. Flood  



### Events that have the greatest economic consequences

```{r}
plot_1 <- ggplot(cropdmg_plot, aes(x=evtype, y=CROPDMG)) + 
  geom_bar(stat = "identity", fill = "blue") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) + 
  ylab("Cost") + ggtitle("Crop") + 
  xlab("Event Type")
plot_2 <- ggplot(propdmg_plot, aes(x=evtype, y=PROPDMG)) + 
  geom_bar(stat = "identity", fill = "green") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) + 
  ylab("Cost") + ggtitle("Property") + 
  xlab("Event Type")
plot_3 <- ggplot(propcropdmg_plot, aes(x=evtype, y=sum)) + 
  geom_bar(stat = "identity", fill = "red") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) + 
  ylab("Cost") + ggtitle("Property + Crop") + 
  xlab("Event Type")
grid.arrange(plot_1, plot_2, plot_3, ncol = 3, main = "Top 10 Events Having the Greatest Economic Consequences")
```



#### The top 3 events that resulted in most crop damages are  
1. Drought
2. Flood
3. River Flood  

#### The top 3 events that resulted in most property damages are
1. Flood
2. Hurricane/Typhoon
3. Storm Surge

#### The top 3 events taking into consideration both crop and property damages that have the greatest economic consequences are
1. Flood
2. Hurricane/Typhoon
3. Storm Surge